GPU-Beschleunigung der Material Point Method

### 2018-11-24 mscthesis.tex
grid nodal forces

\section{Introduction} \label{intro}
bib-test \cite{MPM:SNOW}
$$ -  \int_{\pi/2}^{\pi} \frac{1}{\sqrt{u}} \frac{\lvert \cos{x} \rvert}{\cos{x}} du $$
\subsection{Motivation}
\section{Related Work}
\cite{PIC:GPU} -> \cite{NVIDIA:NNSEARCH}
benefits over radix sort /thrust sort-by-key -> counting sort
\cite{NVIDIA:SHUFFLE}

@@

The first aim will be to find the strain energy density $\psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
For an (hyper-)elastic material the Cauchy stress may be defined in terms of $\psi$ as:

@@

  \item \label{it:mixed_partials}Equivalence of second-order mixed partials of $\psi$:

@@

      C _ { i j k l } = \frac { \partial ^ { 2 } { \psi } } { \partial \epsilon _ { k l } \partial \epsilon _ { i j } } = \frac { \partial ^ { 2 } { \psi } } { \partial \epsilon _ { i j } \partial \epsilon _ { k l } } = C _ { k l i j },

@@

  \sigma _ { i j } = \sigma _ { i j } ( \boldsymbol{\epsilon} ) = \frac { \partial { \psi } } { \partial \epsilon _ { i j } },

@@

if such a $\psi$ is given. An isotropic (direction-independent) linear elastic material further only has three unique elements $C _ {i j k l}$. Using Voigt-Notation, which collapses indices $i=j$ and $k=l$, equation \ref{eq:hook} can be rewritten as:

@@

Comparing with equation \ref{eq:voigt} leads to coefficients $\gamma_{ij} = 2\epsilon_{ij} \text{ for } (i \neq j)$, $C_{11} = \lambda +2 \mu$, $C_{12} = \lambda$ and $C_{22} = \mu$. $\gamma$ is also referred to as the engineering strain. Due to the relationship in \ref{eq:partial_energy} the model for linear elasticity in terms of the strain energy density function $\psi_{LE}$ after integration of \ref{eq:stress_strain} concludes to:

@@

  \psi_{LE} = \mu \Vert \boldsymbol{\epsilon} \Vert^2_F + \frac{\lambda}{2} \text{tr}^2(\boldsymbol{\epsilon}).

@@

\subsubsection{Corotated Hyperelasticity}

@@

The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t. to $\boldsymbol{F}_E$: $\psi(\boldsymbol{\epsilon}(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \psi(\boldsymbol{\epsilon}(\boldsymbol{F}_E))$. Rigid body motions however don't result in strain and consequently don't need to be recovered from. So energy should not change. Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$, an alternate strain measure may be defined as:

@@

  \Psi = \mu \| \boldsymbol{ S }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { S }_E - \boldsymbol { I } )

@@

  \Psi = \mu \| \boldsymbol {F}_E  - \boldsymbol {R}_E \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E - \boldsymbol { I } \right)

@@

  \Psi(\boldsymbol{\Sigma}_E) = \mu \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma }_E - \boldsymbol { I } )

@@

  \boldsymbol{P} = \boldsymbol{\sigma}\boldsymbol{F}^{-T}J

@@

\subsubsection{APIC momentum Transfers}

@@

Motivated by the theory of angular momentum and moment of inertia one can define a quantity
\begin{equation}
\boldsymbol { D } _ { p } ^ { n } = \sum _ { i } w _ { i p } ^ { n } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ^ { T },
\end{equation}
which is similar to the classically known inertia tensor:
\begin{equation}
  \boldsymbol{I} _ { p } = - \sum _ { i } m _ { i } [\boldsymbol{x} _ { i }-\boldsymbol{x}_{p}][\boldsymbol{x} _ { i }-\boldsymbol{x}_{p}]^T
\end{equation}
$$
=
\sum _ { i } m _ { i }((\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})^T(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})\boldsymbol{I} -(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})^T)
$$
Bearing in mind that $\boldsymbol{D}_p^n$ does not include a mass and is defined for an affine motion instead of an angular motion, where $[a]_{\alpha\gamma} = \varepsilon_{\alpha\beta\gamma}a_{\beta}$ is the cross-product matrix and $\boldsymbol{I}$ denotes the identity matrix.
In classical mechanics the angular velocity $\boldsymbol{\omega_p}$ can be then described using the inertia tensor $\boldsymbol{I}_p$ with the help of the angular momentum $\boldsymbol{L}_p$:
\begin{equation}
  \boldsymbol{\omega}_p = \boldsymbol{I}_p^{-1} \boldsymbol{L}_p.
\end{equation}
This motivates the velocity derivative to be equally defined by a similar relationship, where $\boldsymbol{B}^n_p$ holds momentum information.
\begin{equation}
  \boldsymbol{C}^n_p = (\boldsymbol{D}_p^n)^{-1} \boldsymbol{B}^n_p.
\end{equation}
The transfers of the $APIC$-scheme are then summarized:
\begin{enumerate}
\item Particle to grid:
  \begin{equation}\label{eq:apic_mom_p2g}
    (m\boldsymbol{v})^n_i = \sum_p w^n_{ip}m_p (\boldsymbol { v } _ { p } ^ { n } + \boldsymbol { B } _ { p } ^ { n } ( \boldsymbol { D } _ { p } ^ { n } ) ^ { - 1 } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) )
\end{equation}
  \item Factoring out mass.
    \begin{equation}
      \boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}.
    \end{equation}

  \item Grid to Particle Transfer (in a $PIC$-manner):

@@

+ ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } -  { \boldsymbol { x } } _ { i } ^ { n + 1 } + \boldsymbol { x } _ { p } ^ { n + 1 } ) (  { \boldsymbol { v } } _ { i } ^ { n + 1 } ) ^ { T } )
      \boldsymbol{v}_{p}^{n+1} = \sum_i w_{ip}\boldsymbol{v}_i^{n+1}.

@@

@@

Any conservative integrator could be used to discretize time. The implicit midpoint rule ($\lambda =\frac{1}{2}$) is one of such
\begin{equation}\label{eq:midpoint}

@@

Replacing the Lagrangian acceleration $_0a_\alpha$ with the left side of \ref{eq:midpoint} using velocity, taking care of the right side and pushing forward to Eulerian view:

@@

@@

@@

@@

Starting from the last update on the grid which is typically the position with eq. \ref{eq:midpoint}:

@@

\frac{\partial _t\boldsymbol{x}}{\partial t} = {_t\boldsymbol{v}} \Rightarrow \frac{\boldsymbol{\hat{x}}_i^{n+1} - \boldsymbol{x}_i^{n}}{\Delta t} = \boldsymbol{v}_i^{n+\lambda} =  (1-\lambda)\boldsymbol{v}_i^n + \lambda\boldsymbol{\hat{v}}_i^{n+1}

@@

  \Leftrightarrow \boldsymbol{\hat{x}}_i^{n+1}   =\boldsymbol{x}_i^{n} + \Delta t \left((1-\lambda)\boldsymbol{v}_i^n + \lambda\boldsymbol{\hat{v}}_i^{n+1}\right).

@@

Due to \ref{eq:space_discr} a velocity update can be put together as

@@

  \frac{1}{\Delta t} (({m^n\boldsymbol{\hat{v}}^{n+1}})_i-({m^n\boldsymbol{v}^n})_i)
  = \boldsymbol{f}_i^{n+\lambda}
 \Leftrightarrow

@@

 = \boldsymbol{v}^n_i+ \frac{\Delta t}{m_i^n}\boldsymbol{f}_i^{n+\lambda},

@@

where the force can be also described in terms of the in-between point:

@@

 \boldsymbol{f}_i^{n+\lambda} = \boldsymbol{\sigma}^{n+\lambda}_p(\boldsymbol{F}_p^{n+\lambda}(\boldsymbol{\hat{x}}_i^{n+1})).

>>>>

\section{Introduction} \label{intro}
bib-test \cite{MPM:SNOW}
$$ -  \int_{\pi/2}^{\pi} \frac{1}{\sqrt{u}} \frac{\lvert \cos{x} \rvert}{\cos{x}} du $$
\section{Related Work}
\cite{PIC:GPU} -> \cite{NVIDIA:NNSEARCH}
benefits over radix sort /thrust sort-by-key -> counting sort
\cite{NVIDIA:SHUFFLE}
\cite{MPM:POLYPIC}

@@

The first aim will be to find the strain energy density $\Psi(\boldsymbol{\epsilon})$ of the strain $\bm{\epsilon}$.
For an (hyper-)elastic material the Cauchy stress may be defined in terms of $\Psi$ as:

@@

  \item \label{it:mixed_partials}Equivalence of second-order mixed partials of $\Psi$:

@@

      C _ { i j k l } = \frac { \partial ^ { 2 } { \Psi } } { \partial \epsilon _ { k l } \partial \epsilon _ { i j } } = \frac { \partial ^ { 2 } { \Psi } } { \partial \epsilon _ { i j } \partial \epsilon _ { k l } } = C _ { k l i j },

@@

  \sigma _ { i j } = \sigma _ { i j } ( \boldsymbol{\epsilon} ) = \frac { \partial { \Psi } } { \partial \epsilon _ { i j } },

@@

if such a $\Psi$ is given. An isotropic (direction-independent) linear elastic material further only has three unique elements $C _ {i j k l}$. Using Voigt-Notation, which collapses indices $i=j$ and $k=l$, equation \ref{eq:hook} can be rewritten as:

@@

Comparing with equation \ref{eq:voigt} leads to coefficients $\gamma_{ij} = 2\epsilon_{ij} \text{ for } (i \neq j)$, $C_{11} = \lambda +2 \mu$, $C_{12} = \lambda$ and $C_{22} = \mu$. $\gamma$ is also referred to as the engineering strain. Due to the relationship in \ref{eq:partial_energy} the model for linear elasticity in terms of the strain energy density function $\Psi_{LE}$ after integration of \ref{eq:stress_strain} concludes to:

@@

  \Psi_{LE} = \mu \Vert \boldsymbol{\epsilon} \Vert^2_F + \frac{\lambda}{2} \text{tr}^2(\boldsymbol{\epsilon}).

@@

\subsubsection{Corotated Hyperelasticity}\label{sec:cor_hyper}

@@

The energy produced by equation \ref{eq:linear_elas} using the small strain tensor is not rotationally invariant, w.r.t. to $\boldsymbol{F}_E$: $\Psi_{LE}(\boldsymbol{\epsilon}(\boldsymbol{R}_{0} \boldsymbol{F}_E)) \neq \Psi_{LE}(\boldsymbol{\epsilon}(\boldsymbol{F}_E))$. Rigid body motions however don't result in strain and consequently don't need to be recovered from. So energy should not change. Given the polar decomposition $\boldsymbol{F}_E = \boldsymbol{R}_E\boldsymbol{S}_E$, an alternate strain measure may be defined as:

@@

  \Psi_{CH} = \mu \| \boldsymbol{ S }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { S }_E - \boldsymbol { I } )

@@

  \Psi_{CH} = \mu \| \boldsymbol {F}_E  - \boldsymbol {R}_E \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } \left( \boldsymbol { R }_E ^ { T } \boldsymbol { F }_E - \boldsymbol { I } \right)

@@

  \Psi_{CH}(\boldsymbol{\Sigma}_E) = \mu \| \boldsymbol { \Sigma }_E - \boldsymbol { I } \| _ { F } ^ { 2 } + \frac { \lambda } { 2 } \operatorname { tr } ^ { 2 } ( \boldsymbol { \Sigma }_E - \boldsymbol { I } )

@@

  \boldsymbol{P} =
  \frac{\partial \Psi}{\partial \boldsymbol{F}}=
  \boldsymbol{\sigma}\boldsymbol{F}^{-T}J.

@@

\subsubsection{APIC transfers}

@@

Motivated by the theory of angular momentum and moment of inertia one can define a quantity
\begin{equation}
\boldsymbol { D } _ { p } ^ { n } = \sum _ { i } w _ { i p } ^ { n } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) ^ { T },
\end{equation}
which is similar to the classically known inertia tensor:
\begin{equation}
  \boldsymbol{I} _ { p } = - \sum _ { i } m _ { i } [\boldsymbol{x} _ { i }-\boldsymbol{x}_{p}][\boldsymbol{x} _ { i }-\boldsymbol{x}_{p}]^T
\end{equation}
$$
=
\sum _ { i } m _ { i }((\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})^T(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})\boldsymbol{I} -(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})(\boldsymbol{x} _ { i }-\boldsymbol{x}_{p})^T)
$$
Bearing in mind that $\boldsymbol{D}_p^n$ does not include a mass and is defined for an affine motion instead of an angular motion, where $[a]_{\alpha\gamma} = \varepsilon_{\alpha\beta\gamma}a_{\beta}$ is the cross-product matrix and $\boldsymbol{I}$ denotes the identity matrix.
In classical mechanics the angular velocity $\boldsymbol{\omega_p}$ can be then described using the inertia tensor $\boldsymbol{I}_p$ with the help of the angular momentum $\boldsymbol{L}_p$:
\begin{equation}
  \boldsymbol{\omega}_p = \boldsymbol{I}_p^{-1} \boldsymbol{L}_p.
\end{equation}
This motivates the velocity derivative to be equally defined by a similar relationship, where $\boldsymbol{B}^n_p$ holds momentum information.
\begin{equation}
  \boldsymbol{C}^n_p = (\boldsymbol{D}_p^n)^{-1} \boldsymbol{B}^n_p.
\end{equation}
The transfers of the $APIC$-scheme are then summarized:
\begin{enumerate}
\item Particle to grid:
  \begin{equation}\label{eq:apic_mom_p2g}
    (m\boldsymbol{v})^n_i = \sum_p w^n_{ip}m_p (\boldsymbol { v } _ { p } ^ { n } + \boldsymbol { B } _ { p } ^ { n } ( \boldsymbol { D } _ { p } ^ { n } ) ^ { - 1 } ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } ) )
\end{equation}
  \item Factoring out mass:
    \begin{equation}
      \boldsymbol{v}_i^n = \frac{(m\boldsymbol{v})_i^n}{m_i^n}.
    \end{equation}

  \item Grid to particle transfer (in a $PIC$-manner), where in contrast the new particle position $\boldsymbol{x}_p$ also needs to be interpolated:
    \begin{equation}
      \boldsymbol{x}_{p}^{n+1} = \sum_i w_{ip}\boldsymbol{x}_i^{n+1}
    \end{equation}
      \boldsymbol{v}_{p}^{n+1} = \sum_i w_{ip}\boldsymbol{v}_i^{n+1}

@@

+ ( \boldsymbol { x } _ { i } ^ { n } - \boldsymbol { x } _ { p } ^ { n } -  { \boldsymbol { x } } _ { i } ^ { n + 1 } + \boldsymbol { x } _ { p } ^ { n + 1 } ) (  { \boldsymbol { v } } _ { i } ^ { n + 1 } ) ^ { T } ).

@@

\begin{equation} \label{eq:weak_mpm}

@@

Any integrator conserving linear and angular momentum could be used to discretize time. The class of time integrators used here are characterized by \begin{equation}\label{eq:midpoint}

@@

A prominent member of this class is the implicit midpoint rule ($\lambda =\frac{1}{2}$). Replacing the Lagrangian acceleration $_0a_\alpha$ in eq. \ref{eq:weak_mpm} with the left side of \ref{eq:midpoint} using velocity, taking care of the right side and pushing forward to Eulerian view:

@@

\begin{equation}\label{eq:force_disc}

@@

\begin{equation}\label{eq:evol_def_grad_disc}

@@

The discretization of the position will also be of need to advance the particles and weight them back:
\begin{equation}  \frac{\partial _t\boldsymbol{x}}{\partial t} = {_t\boldsymbol{v}} \Rightarrow \frac{\boldsymbol{\hat{x}}_i^{n+1}
  - \boldsymbol{x}_i^{n}}{\Delta t} = \boldsymbol{v}_i^{n+\lambda}.
The grid position $\boldsymbol{\hat{x}}_i^{n+1}$ does not correspond to an actual deformation. The grid never actually gets deformed (unlike in FEM-methods). Therefore the discretized evolution of the deformation gradient  \ref{eq:evol_def_grad_disc} is directly a function of $\boldsymbol{\hat{x}}$. For the point $\boldsymbol{\hat{x}}_i^{n+1}$ this becomes:
\begin{equation}\label{eq:evol_def_grad_disc_x}
  \boldsymbol{\hat{F}}^{n+1}_p(\boldsymbol{\hat{x}}_i^{n+1}) =  \left( \boldsymbol{I} + \sum_i (\boldsymbol{\hat{x}}_i^{n+1} - \boldsymbol{x}_i^{n})({\nabla w_{ip}})^T \right)\boldsymbol{F}^{n}_p.
As part of the class of time integrators in use (\ref{eq:midpoint}) a function of $\boldsymbol{x}_i$ gets evaluated at an in-between point given by:
  \boldsymbol{x}_i^{n+\lambda} = \lambda \boldsymbol{x}_i^{n+1} + (1-\lambda)\boldsymbol{x}_i^n.
Plugging this point into \ref{eq:evol_def_grad_disc_x} leads to the following generalization:
$$\boldsymbol{\hat{F}}^{n+\lambda}_p(\boldsymbol{x}_i^{n+\lambda}) = \left( \boldsymbol{I} + \lambda \sum_i (\boldsymbol{\hat{x}}_i^{n+1} - \boldsymbol{x}_i^{n})({\nabla w_{ip}})^T \right)\boldsymbol{F}^{n}_p
  \begin{equation} \label{eq:def_grad_lambda}
  = (1-\lambda)\boldsymbol{F}^n_p + \lambda \boldsymbol{F}^{n+1}_p
This

@@

The notion of a total elastic potential energy function $\Psi$ was introduced in chapter \ref{sec:cor_hyper}. The MPM approximation to this function can be defined by:
  e(\hat{\boldsymbol{x}}) = \sum_pV^0_p \Psi(\hat{\boldsymbol{F}}_{Ep}(\hat{\boldsymbol{x}})).
The evolution of the deformation gradient \ref{eq:evol_def_grad_disc_x} for a general $\boldsymbol{\hat{x}}_i$  is necessary
$$\frac{\partial \hat{F}_{\omega\beta}}{\partial \hat{x}_\alpha}
= \delta_{\omega \alpha}
\frac{\partial w_{ip}}{\partial x_\gamma}F_{\gamma\beta} = \delta_{\omega \alpha}F_{\gamma\beta} \frac{\partial{w_{ip}}}{\partial x_\gamma}$$
for the spatial derivative of the potential $e(\boldsymbol{\hat{x}})$. This is just the force created by elastic stresses out of eq. \ref{eq:force_disc}:

@@

  \frac{\partial e}{\partial \hat{x}_{i\alpha}}(\hat{\boldsymbol{x}})
  = \sum_p V^0_p \frac{\partial \Psi}{\partial \hat{F}_{\omega\beta}}(\hat{\boldsymbol{F}}_{Ep})\frac{\partial \hat{F}_{\omega\beta}}{\partial \hat{x}_\alpha}

@@

  = \sum_p V^0_p P_{\alpha\beta}(\hat{\boldsymbol{F}}_{Ep})F_{\gamma\beta}\frac{\partial w_{ip}}{\partial x_\gamma} = -\hat{f}_{i\alpha}

@@

where the force can be also described in terms of the Cauchy stress:

@@

  \boldsymbol{\hat{f}}_i = - \sum_p V^n_p \boldsymbol{\sigma}_p(\boldsymbol{\hat{F}}_{Ep})\nabla{w^n_{ip}}.
Due to eq. \ref{eq:def_grad_lambda} we can for the class of functions summarize the stress computation as:
  \boldsymbol{P}^{n+\lambda}(\hat{\boldsymbol{F}}_{Ep}) = \boldsymbol{P}(\hat{\boldsymbol{F}}^{n+\lambda}_{Ep}),
  \quad
  \boldsymbol{\sigma}^{n+\lambda}(\hat{\boldsymbol{F}}_{Ep}) = \boldsymbol{\sigma}(\hat{\boldsymbol{F}}^{n+\lambda}_{Ep}).
\cite{MPM:COURSE}\cite{MPM:APIC}
\subsubsection{Symplectic midpoint scheme}
Starting from the last update on the grid which is typically the position with eq. \ref{eq:midpoint} the general midpoint scheme ($\lambda=\frac{1}{2}$) is:
  \boldsymbol{\hat{x}}_i^{n+1}   =\boldsymbol{x}_i^{n} + \Delta t \boldsymbol{v}_i^{n+\frac{1}{2}}.
Due to \ref{eq:space_discr} a velocity update can be put together as:

@@

 = \boldsymbol{v}^n_i+ \frac{\Delta t}{m_i^n}\boldsymbol{f}_i^{n+\frac{1}{2}}.

@@

@@
